% SoftwareX "Original Software Publication" metapaper.
% Uses the Elsevier elsarticle class (ships with TeX Live's elsarticle package).
\documentclass[preprint,12pt]{elsarticle}

\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{url}

% Placeholder marker: items that require an author decision or a future run are wrapped in \FILL and
% rendered in red, so an unfilled template can never be mistaken for a finalized value.
\newcommand{\FILL}[1]{{\color{red}\textbf{[FILL: #1]}}}

\journal{SoftwareX}

\begin{document}

\begin{frontmatter}

\title{Leviathan: An Open, Reproducible Pipeline for Statistical Text Analysis of Judicial Opinions}

\author[ind]{Marcus Castillo}
\ead{marcus.f.castillo@gmail.com}
\address[ind]{\FILL{affiliation / ``Independent Researcher''; add ORCID}}

\begin{abstract}
We present \textbf{Leviathan}, an open-source, containerized pipeline for the statistical analysis of
judicial opinions. Leviathan ingests opinions and metadata, segments U.S. Supreme Court opinions into
majority, concurrence, and dissent components, and computes descriptive, lexical, embedding-, and
graph-based measures: outcome and tone statistics, majority-versus-dissent lexical divergence via the
weighted log-odds-ratio with an informative Dirichlet prior, topic models, per-justice writing-style
embeddings, and citation-network influence and community structure. The system is built around an
explicit \emph{ethics-first} design: every figure is emitted with a confidence score, a sample-size
warning, and a limitations statement, and the software is constrained to report statistical
disparities only---never claims of intent, ideology, or bias. We describe the architecture and
methods, demonstrate the pipeline end-to-end on real Supreme Court opinions, and provide a
gold-label validation harness against the Supreme Court Database (SCDB) that we verify against the
complete 2024 release (8{,}822 citable cases; 98\% of issue areas covered by our crosswalk). All
code, example data, dataset-versioning tooling, a reproduction recipe, and continuous-integration
configuration are publicly available.
\end{abstract}

\begin{keyword}
legal NLP \sep judicial opinions \sep reproducibility \sep text-as-data \sep Supreme Court Database
\sep research software
\end{keyword}

\end{frontmatter}

% ----------------------------------------------------------------------------- %
\section*{Required Metadata}
\section*{Code metadata}

\begin{table}[h!]
\centering
\small
\begin{tabular}{|p{0.31\textwidth}|p{0.42\textwidth}|p{0.17\textwidth}|}
\hline
\textbf{Nr.} & \textbf{Code metadata description} & \textbf{Value} \\ \hline
C1 & Current code version & v0.1.0 \\ \hline
C2 & Permanent link to code/repository & \url{https://github.com/marcus-castillo/Leviathan} (archive: \FILL{Zenodo DOI}) \\ \hline
C3 & Permanent link to Reproducible Capsule & \FILL{Code Ocean / none} \\ \hline
C4 & Legal Code License & MIT \\ \hline
C5 & Code versioning system used & git \\ \hline
C6 & Software code languages, tools, services used & Python, TypeScript; FastAPI, Next.js; PostgreSQL/pgvector, Neo4j; scikit-learn, sentence-transformers, spaCy; Docker \\ \hline
C7 & Compilation requirements, operating environments \& dependencies & Docker Compose; Python 3.11; Node 20; see \texttt{requirements.txt} \\ \hline
C8 & Link to developer documentation/manual & Repository \texttt{README.md} files and \texttt{paper/REPRODUCE.md} \\ \hline
C9 & Support email for questions & marcus.f.castillo@gmail.com \\ \hline
\end{tabular}
\caption{Code metadata.}
\end{table}

% ----------------------------------------------------------------------------- %
\section{Motivation and significance}
Empirical study of judicial behavior increasingly draws on natural language processing, yet much of
this work relies on bespoke, unpublished pipelines whose preprocessing, labeling, and aggregation
choices are difficult to inspect or reproduce. Two needs follow. First, the field needs
\emph{shared, auditable infrastructure}: a documented path from raw opinions to analysis-ready
measures, with versioned datasets and automated tests. Second---because measures framed around
``judicial bias'' are ethically fraught and easily over-interpreted---the field needs tooling whose
design makes over-claiming structurally difficult.

Leviathan addresses both. It is a four-component system---(i) a bias-metrics API and dashboard,
(ii) a dataset-standardization command-line tool with semantic versioning and reproducibility logs,
(iii) a Neo4j citation-graph service with influence and community analysis, and (iv) a Supreme Court
analysis platform---sharing a common embedding space and a common ethics layer.

Quantitative analysis of opinions spans political science and legal NLP. The Supreme Court Database
\cite{spaeth2023scdb} supplies authoritative, human-coded case-level labels (winning party, decision
direction, issue area) and is the standard ground truth for work on the U.S. Supreme Court;
ideal-point models such as Martin--Quinn scores \cite{martin2002dynamic} estimate ideological
position from votes. Distinctive-vocabulary analysis follows Monroe, Colaresi, and Quinn
\cite{monroe2008fightin}; text representation builds on sentence embeddings
\cite{reimers2019sentencebert} and domain language models \cite{chalkidis2020legalbert}, with
legal-outcome prediction studied since at least \cite{aletras2016predicting}. Leviathan's
contribution is not a new algorithm but the \emph{integration, validation tooling, and
ethics-constrained packaging} of these established methods into reusable research software, together
with a verified gold-label harness and an exact reproduction recipe.

\paragraph{Scope.} This is a software/resource contribution. We do not claim to detect bias, intent,
or ideology; we provide a reproducible pipeline, a validation harness against authoritative human
codings, and a design that enforces caveated, disparity-only reporting.

% ----------------------------------------------------------------------------- %
\section{Software description}

\subsection{Software architecture}
Leviathan is a polyglot monorepo orchestrated with Docker Compose. A FastAPI backend exposes the
bias-metrics API; a PostgreSQL instance with the \texttt{pgvector} extension stores opinions,
analyses, and 384-dimensional sentence embeddings (\texttt{all-MiniLM-L6-v2}). The \texttt{levcorpus}
command-line tool produces standardized JSONL/CSV/Parquet datasets with a semantic-version registry
and reproducibility logs. A Neo4j service stores the citation graph; influence, clustering, and
label-neutral statistical-grouping analytics run on NetworkX projections so they are unit-testable
without a live database. The SCOTUS service adds opinion segmentation and per-justice analysis. Three
Next.js dashboards and a static landing page provide visualization. Continuous integration runs all
Python test suites, byte-compilation, and frontend type-checks on every push.

\subsection{Software functionalities}
\begin{itemize}
  \item \textbf{Ingestion} of CourtListener \cite{freelaw} (REST and bulk) and plain-text opinions
        with metadata, with idempotent upserts and judge/court normalization.
  \item \textbf{Opinion segmentation} into majority, concurrence, dissent, and per-curiam parts. For
        CourtListener data we use the publisher's typed sub-opinions; for raw text we apply an
        auditable authorship/role-header detector with authoring-justice attribution.
  \item \textbf{Lexical divergence} via the weighted log-odds-ratio with an informative Dirichlet
        prior \cite{monroe2008fightin}. For word $w$ with counts $y^{(i)}_w$ in corpus
        $i\in\{A,B\}$, totals $n^{(i)}$, and a prior $\alpha_w$ derived from the combined corpus,
        \begin{equation}
        \delta_w = \log\!\frac{y^{(A)}_w+\alpha_w}{n^{(A)}+\alpha_0-y^{(A)}_w-\alpha_w}
                 - \log\!\frac{y^{(B)}_w+\alpha_w}{n^{(B)}+\alpha_0-y^{(B)}_w-\alpha_w},
        \quad
        z_w = \frac{\delta_w}{\sqrt{1/(y^{(A)}_w+\alpha_w)+1/(y^{(B)}_w+\alpha_w)}},
        \end{equation}
        where $\alpha_0=\sum_w \alpha_w$. The prior shrinks rare-word noise that raw frequency and
        tf--idf do not. We apply this to majority-versus-dissent text and to per-justice ``framing''.
  \item \textbf{Topic modeling} via non-negative matrix factorization \cite{lee1999nmf} over tf--idf,
        plus a curated constitutional-theme lexicon whose tags trace to matched terms.
  \item \textbf{Justice style embeddings}: mean-pooled sentence embeddings
        \cite{reimers2019sentencebert} yielding pairwise similarity and $k$-means style clusters with
        neutral (non-ideological) labels.
  \item \textbf{Citation-graph analytics}: PageRank, betweenness, and Louvain communities
        \cite{blondel2008louvain} over typed citation edges.
  \item \textbf{Outcome classification}: a transparent rule-based disposition classifier, and a
        trainable TF--IDF + logistic-regression model with a held-out evaluation harness against SCDB
        \texttt{partyWinning}.
  \item \textbf{Ethics layer}: every emitted signal carries a confidence score, a sample-size warning
        below a configurable threshold, and a limitations string; narrative text is programmatically
        scrubbed of intent/ideology vocabulary.
\end{itemize}
Built with spaCy \cite{honnibal2020spacy} and scikit-learn \cite{pedregosa2011scikit}; open corpora
from CourtListener \cite{freelaw} and the Caselaw Access Project \cite{cap}.

% ----------------------------------------------------------------------------- %
\section{Illustrative examples}

\subsection{End-to-end on a real opinion}
We ran the pipeline on real Supreme Court opinions retrieved from CourtListener and joined to SCDB.
As a concrete example, for \emph{Arave v.\ Hoffman} (OT2007), Leviathan's theme tagger assigns
\texttt{criminal-procedure}, and the weighted log-odds analysis of the majority opinion against the
rest of the sample surfaces the most distinctive terms \texttt{plea}, \texttt{ineffective},
\texttt{assistance}, \texttt{counsel}, and \texttt{dismiss}---correctly characterizing an
ineffective-assistance-of-counsel matter. Such per-case outputs are produced by the
\texttt{/analysis/*} and \texttt{/justice/*} endpoints and rendered in the SCOTUS dashboard.

\subsection{Gold-label validation harness against SCDB}
\label{sec:validation}
Leviathan ships a validation harness (\texttt{scotusapp/validation}) that joins ingested opinions to
SCDB \cite{spaeth2023scdb} by SCDB \texttt{caseId} (CourtListener's \texttt{scdb\_id}), falling back
to normalized U.S.\ citation, and reports agreement of the label-producing components with the human
codings: \emph{issue/theme} (our lexicon tagger vs.\ SCDB \texttt{issueArea}, over a documented
crosswalk; ambiguous areas excluded) and \emph{outcome} (a TF--IDF + logistic-regression classifier
trained on SCDB \texttt{partyWinning} with a stratified held-out split, compared against the
rule-based baseline). Metrics are accuracy, macro-$F_1$, and Cohen's $\kappa$.

We verified the harness against the complete SCDB 2024 release: 8{,}822 citable cases load and index
by \texttt{caseId}, and 98\% fall in an issue area covered by our crosswalk. The end-to-end retrieval
and scoring pipeline executed successfully on real opinions. We do not report population-scale
agreement statistics here: the live retrieval used for this manuscript was constrained by the data
provider's API rate limits, yielding a sample too small for inference. The full-scale validation
(target $N\!\approx\!500$, terms 2003--2019) is specified end-to-end in \texttt{paper/REPRODUCE.md}
and runs unattended from the bulk corpus; we therefore present the harness and methodology here and
report the population metrics in Table~\ref{tab:val} once the full run completes.

\begin{table}[h!]
\centering
\begin{tabular}{lrrrr}
\toprule
Component & $N$ & Accuracy & Macro-$F_1$ & Cohen's $\kappa$ \\
\midrule
Issue/theme vs.\ SCDB \texttt{issueArea} & \FILL{N} & \FILL{acc} & \FILL{F1} & \FILL{$\kappa$} \\
Outcome (trained) vs.\ SCDB \texttt{partyWinning} & \FILL{N} & \FILL{acc} & \FILL{F1} & \FILL{$\kappa$} \\
Outcome (rule-based baseline) & \FILL{N} & \FILL{acc} & --- & --- \\
\bottomrule
\end{tabular}
\caption{Agreement of Leviathan component outputs with SCDB gold codings, from the full-scale run
(corpus: \FILL{terms, source, SCDB release}). Produced by \texttt{scripts/train\_eval\_outcome.py};
not hand-entered.}
\label{tab:val}
\end{table}

% ----------------------------------------------------------------------------- %
\section{Impact}
Leviathan lowers the barrier to transparent, reproducible empirical work on judicial text. By
packaging ingestion, segmentation, validated labeling, embeddings, and graph analytics behind a
common API, a dataset-versioning tool, and a tested codebase, it lets researchers reproduce and
extend analyses rather than re-implement pipelines. The SCDB validation harness gives users a direct,
authoritative yardstick for the noisy components and a one-command path to recompute agreement on
their own corpus. The ethics layer encodes responsible-reporting norms directly in software, which is
useful for teaching and for oversight-oriented research where over-claiming carries real cost. The
dataset standardizer additionally produces analysis-ready, version-stamped corpora reusable beyond
this system.

% ----------------------------------------------------------------------------- %
\section{Limitations}
Outcome, tone, and theme labels are model-generated and noisy; the rule-based outcome classifier in
particular under-fires on Supreme Court prose, motivating the trainable model. Corpus coverage is
uneven across courts and eras, so absence of data is not absence of activity. Embedding
``similarity'' need not imply legal equivalence. Statistical groupings are artifacts of the corpus,
feature set, and parameters and are \emph{not} ideology measures---for ideology, validated external
instruments such as Martin--Quinn \cite{martin2002dynamic} should be used. Nothing in the system
supports causal claims about any judge. Finally, large-scale public retrieval is bounded by data
providers' API rate limits; reproducible full-scale validation is therefore documented against the
bulk corpus rather than the live API, and the bundled example corpora are synthetic, for
smoke-testing only.

% ----------------------------------------------------------------------------- %
\section{Conclusions}
Leviathan packages established legal-NLP and graph methods into an auditable, reproducible, and
ethically constrained pipeline, with a gold-label validation harness verified against the full
Supreme Court Database. We hope it lowers the barrier to transparent empirical work on judicial text
while making over-claiming structurally hard.

\section*{Acknowledgements}
\FILL{funding / acknowledgements, and disclosure of AI-assisted development per SoftwareX policy.}

\bibliographystyle{elsarticle-num}
\bibliography{references}

\end{document}
